Diffeomorphic Directly Manipulated Free-Form Deformation Image Registration via Vector Field Flows

### 2012-01-28 ANTsTBSS/paperRevision.tex
ENH: More stuff.

Thus, instead of normalization based on anatomical alignment followed by
statistical testing, direct normalization of the images to be statistically compared using SSD
explicitly conflates
anatomical alignment with optimizing the very statistical testing
results that one will ultimately use to assess hypotheses.  This immediately evokes a sense of circularity in the analysis \citep{Kriegeskorte2010}.
According to Kriegeskrote et al., ``An analysis is circular (or nonindependent) if it is based on data that were selected for showing the effect of interest or a related effect.''  Direct FA-to-FA template registration will produce transformations that align (or select) voxels such that the statistical testing result (or the effect of interest) is increased (assuming one is using Student's 
$t$-test).  While the SSD metric is able to provide visually attractive alignments across subjects, the voxel-level details (that is, which voxels are matched where) are strikingly important in terms of the effect on detection power.

>>>>

Thus, instead of normalization based on anatomical alignment followed by
statistical testing, direct normalization of the images to be statistically compared using SSD
explicitly conflates {\color{red}{attempts at}}
anatomical alignment with optimizing the very statistical testing
results that one will ultimately use to assess hypotheses.  This immediately evokes a sense of circularity in the analysis \citep{Kriegeskorte2010}.
According to Kriegeskrote et al., ``An analysis is circular (or nonindependent) if it is based on data that were selected for showing the effect of interest or a related effect.''  Direct FA-to-FA template registration will produce transformations that align (or select) voxels such that the statistical testing result (or the effect of interest) is increased (assuming one is using Student's 
$t$-test).  While the SSD metric is able to provide visually attractive alignments across subjects, the voxel-level details (that is, which voxels are matched where) are strikingly important in terms of the effect on detection power.

### 2012-01-28 WBIR2012/paper.tex
ENH: More stuff.

% This is LLNCS.DEM the demonstration file of
% the LaTeX macro package from Springer-Verlag
% for Lecture Notes in Computer Science,
% version 2.4 for LaTeX2e as of 16. April 2010
%
\documentclass{llncs}
%
\usepackage{makeidx}  % allows for indexgeneration
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{algorithmic,eqparbox,array}
\usepackage{algorithm}

@@

\begin{abstract}
Motivated by previous work \cite{tustison2009} and 
recent diffeomorphic image registration developments in which the 
characteristic velocity field is represented by spatiotemporal B-splines 
\cite{de-craene2011}, we present a diffeomorphic B-spline-based
image registration algorithm combining and extending these techniques.  
The advancements of the proposed framework
over previous work include a preconditioned gradient descent algorithm, 
potential spatial weighting 
of the gradient permitting (among other things) enforcement of 
stationary boundary conditions, multiresolution in terms of the B-spline
mesh, facilitating the incorporation of landmark information, 
and the possibility of modeling temporal periodicity.  In addition to
theoretical and practical discussions of our contribution, we also describe 
its parallelized implementation as open source in the Insight Toolkit.  
\keywords{B-splines, DMFFD, diffeomorphism, ITK}
\end{abstract}

@@

We first describe the methodology by laying out a mathematical description 
of the various algorithmic elements coupled with implementation details
where appropriate.  This is followed by several registration examples
to provide insight into the details of our contribution.

@@

In \cite{tustison2009} 
it was observed that optimization of FFD registration with gradient descent is 
intrinsically susceptible to problematic energy topographies.  However, a
well-understood preconditioned gradient was proposed based on the work 
described in \cite{tustison2006} which substantially 
improves performance which we refer to as DMFFD image registration.  
Similarly, we propose the following velocity field control
point lattice preconditioned gradient,
$\delta v_{i_1,\ldots,i_d,i_t}$, given the similarity metric, $\Pi$,%
\footnote
{
Current options include neighborhood cross correlation (CC), mutual information (MI), and
Demons-style sum of squared differences (Demons).
}
\begin{align}
\label{eq:dmffd}
  \delta v_{i_1,\ldots,i_d,i_t} &= \left( \sum_{c=1}^{N_{\Omega} \times N_t} \left( \frac{\partial \Pi}{\partial \mathbf{x}} \right)_c B_{i_t}(t^c)\prod_{j=1}^d B_{i_j}(x_j^c)  \right. \nonumber \\
  &\cdot \left. \frac{B_{i_t}^2(t^c) \prod_{j=1}^d B_{i_j}^2 (x_j^c)} 
  {\sum_{k_1=1}^{r+1}\ldots\sum_{k_d=1}^{r+1} \sum_{k_t=1}^{r+1} B_{k_t}^2(t^c) 
  \prod_{j=1}^d B_{k_j}^2 (x_j^c)} \right) \nonumber \\
  &\cdot\left({\sum_{c=1}^{N_{\Omega}\times N_t}B_{i_t}^2(t^c) \prod_{j=1}^d B_{i_j}^2 (x_j^c)} \right) ^{-1}
\end{align}
which is a slight modification of Eqn. (21)
in \cite{tustison2009} which takes into account the temporal locations of the 
dense gradient field sampled in $t \in [0,1]$. $N_t$ and $N_\Omega$ are the number
of time point samples and the number of voxels in the reference image domain, respectively.
$r$ is the spline order in all dimensions%
\footnote{
Spline orders can be specified separately for each dimension but, for simplicity,
we only specify a single spline order.  However, the framework can easily 
accommodate different spline orders.
}
and $c$ indexes the spatio-temporal dense metric
gradient sample.
Combining DMFFD optimization with the B-spline velocity field transform discussed in
the previous subsection, we outline the iterative routine for optimizing the 
velocity field control in Algorithm \ref{alg1}.

@@

As mentioned previously, the registration algorithm has been implemented in the
Insight Toolkit and consists of the following major classes: 
\begin{itemize}
  \item \verb#itk::TimeVaryingVelocityFieldIntegrationImageFilter#
  \item \verb#itk::TimeVaryingBSplineVelocityFieldTransform#
  \item \verb#itk::TimeVaryingBSplineVelocityFieldImageRegistrationMethod#
  \item \verb#itk::TimeVaryingBSplineVelocityFieldTransformParametersAdaptor#
  \item \verb#itk::BSplineScatteredDataPointSetToImageFilter#
\end{itemize}
The \verb#TimeVaryingVelocityFieldIntegrationImageFilter# class implements
the Runge-Kutta integration described earlier.  Given a 
velocity field control point lattice as 
input and the lower and upper integration limits, integration is performed 
in a multi-threaded fashion (since each point in the domain
can be integrated separately).  The \verb#TimeVaryingBSplineVelocityFieldTransform# 
is derived from the base transform class which handles the mapping of geometric primitives
for warping images.  The DMFFD gradient calculation  is handled by the class \verb#BSplineScatteredDataPointSetToImageFilter# (cf. Eqn. \ref{eq:dmffd} and line 11
of Algorithm \ref{alg1}) which takes as input the dense similarity metric and an optional
weighting for each gradient sample.  This permits enforcement of stationary physical
boundaries by specifying a zero gradient on the boundaries and a large weighting.
Coordinating all the elements of image registration is the
\verb#TimeVaryingBSplineVelocityFieldImageRegistrationMethod# class which
encapsulates Algorithm \ref{alg1}.  The resolution scheduling is determined
by the \verb#TimeVaryingBSplineVelocityFieldTransformParametersAdaptor# class.
We provide access to the new ITK 
registration framework (including the B-spline velocity field transform) 
through the command line module \verb#antsRegistration# available both in ANTs%
\footnote{
http://www.picsl.upenn.edu/ANTs
}
and accompanied by a technical report offered through the Insight Journal
\cite{tustison2012}.%
\footnote{
http://www.insight-journal.org
}
The interested reader should consult \cite{tustison2012} for further details on 
actual usage.

\section{Registration Examples}

>>>>

% This is LLNCS.DEM the demonstration file of
% the LaTeX macro package from Springer-Verlag
% for Lecture Notes in Computer Science,
% version 2.4 for LaTeX2e as of 16. April 2010
%
\documentclass{llncs}
%
\usepackage{makeidx}  % allows for indexgeneration
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{algorithmic,eqparbox,array}
\usepackage{algorithm}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{color}

    \definecolor{listcomment}{rgb}{0.0,0.5,0.0}
    \definecolor{listkeyword}{rgb}{0.0,0.0,0.5}
    \definecolor{listnumbers}{gray}{0.65}
    \definecolor{listlightgray}{gray}{0.955}
    \definecolor{listwhite}{gray}{1.0}

@@

\begin{abstract}
Motivated by previous work \cite{tustison2009} and 
recent diffeomorphic image registration developments in which the 
characteristic velocity field is represented by spatiotemporal B-splines 
\cite{de-craene2011}, we present a diffeomorphic B-spline-based
image registration algorithm combining and extending these techniques.  
The advancements of the proposed framework
over previous work include a preconditioned gradient descent algorithm, 
potential spatial weighting 
of the gradient permitting (among other things) enforcement of 
stationary boundary conditions, multiresolution in terms of the B-spline
mesh, facilitating the incorporation of landmark information, 
and the possibility of modeling temporal periodicity.  In addition to
theoretical and practical discussions of our contribution, we also describe 
its parallelized implementation as open source in the Insight Toolkit.  
\keywords{B-splines, DMFFD, diffeomorphisms, image registration, ITK}
\end{abstract}

@@

We first describe the methodology by laying out a mathematical description 
of the various algorithmic elements coupled with implementation details
where appropriate.  This is followed by evaluation on a public data set.

@@

In \cite{tustison2009} 
it was observed that optimization of FFD registration with gradient descent is 
intrinsically susceptible to problematic energy topographies.  However, a
well-understood preconditioned gradient was proposed based on the work 
described in \cite{tustison2006} which substantially 
improves performance which we refer to as DMFFD image registration.  
Similarly, we propose the following velocity field control
point lattice preconditioned gradient,
$\delta v_{i_1,\ldots,i_d,i_t}$, given the similarity metric, $\Pi$,%
\footnote
{
Current options include neighborhood cross correlation (CC), mutual information (MI), and
Demons-style sum of squared differences (Demons).
}
\begin{align}
\label{eq:dmffd}
  \delta v_{i_1,\ldots,i_d,i_t} &= \left( \sum_{c=1}^{N_{\Omega} \times N_t} \left( \frac{\partial \Pi}{\partial \mathbf{x}} \right)_c B_{i_t}(t^c)\prod_{j=1}^d B_{i_j}(x_j^c)  \right. \nonumber \\
  &\cdot \left. \frac{B_{i_t}^2(t^c) \prod_{j=1}^d B_{i_j}^2 (x_j^c)} 
  {\sum_{k_1=1}^{r+1}\ldots\sum_{k_d=1}^{r+1} \sum_{k_t=1}^{r+1} B_{k_t}^2(t^c) 
  \prod_{j=1}^d B_{k_j}^2 (x_j^c)} \right) \nonumber \\
  &\cdot\left({\sum_{c=1}^{N_{\Omega}\times N_t}B_{i_t}^2(t^c) \prod_{j=1}^d B_{i_j}^2 (x_j^c)} \right) ^{-1}
\end{align}
which is a slight modification of Eqn. (21)
in \cite{tustison2009} which takes into account the temporal locations of the 
dense gradient field sampled in $t \in [0,1]$. $N_t$ and $N_\Omega$ are the number
of time point samples and the number of voxels in the reference image domain, respectively.
$r$ is the spline order in all dimensions%
\footnote{
Spline orders can be specified separately for each dimension but, for simplicity,
we only specify a single spline order.  
}
and $c$ indexes the spatio-temporal dense metric
gradient sample.
Combining DMFFD optimization with the B-spline velocity field transform discussed in
the previous subsection, we outline the iterative routine for optimizing the 
velocity field control in Algorithm \ref{alg1}.

@@

As mentioned previously, the registration algorithm has been implemented in the
Insight Toolkit and consists of the following major classes: 
\begin{itemize}
  \item \verb#itk::TimeVaryingVelocityFieldIntegrationImageFilter#
  \item \verb#itk::TimeVaryingBSplineVelocityFieldTransform#
  \item \verb#itk::TimeVaryingBSplineVelocityFieldImageRegistrationMethod#
  \item \verb#itk::TimeVaryingBSplineVelocityFieldTransformParametersAdaptor#
  \item \verb#itk::BSplineScatteredDataPointSetToImageFilter#
\end{itemize}
The \verb#TimeVaryingVelocityFieldIntegrationImageFilter# class implements
the Runge-Kutta integration described earlier.  Given a 
velocity field control point lattice as 
input and the lower and upper integration limits, integration is performed 
in a multi-threaded fashion (since each point in the domain
can be integrated separately).  The \verb#TimeVaryingBSplineVelocityFieldTransform# 
is derived from the base transform class which handles the mapping of geometric primitives
for warping images.  The DMFFD gradient calculation  is handled by the class \verb#BSplineScatteredDataPointSetToImageFilter# (cf. Eqn. \ref{eq:dmffd} and line 11
of Algorithm \ref{alg1}) which takes as input the dense similarity metric and an optional
weighting for each gradient sample.  This permits enforcement of stationary physical
boundaries by specifying a zero gradient on the boundaries and a large weighting.
Coordinating all the elements of image registration is the
\verb#TimeVaryingBSplineVelocityFieldImageRegistrationMethod# class which
encapsulates Algorithm \ref{alg1}.  The resolution scheduling is determined
by the \verb#TimeVaryingBSplineVelocityFieldTransformParametersAdaptor# class.
We provide access to the new ITK 
registration framework (including the B-spline velocity field transform) 
through the command line module \verb#antsRegistration# available both in ANTs%
\footnote{
http://www.picsl.upenn.edu/ANTs
}
and accompanied by a technical report offered through the Insight Journal
\cite{tustison2012}.%
\footnote{
http://hdl.handle.net/10380/3334
}
The interested reader should consult \cite{tustison2012} for further details on 
actual usage.

\section{LPBA40 Evaluation}

As a preliminary evaluation for our algorithm we used a portion 
of the approach of Klein et al. \cite{Klein2009} in which 
various open brain data sets were used to evaluate different
segmentation algorithms.  Specifically, we used the first 20 
subjects of the LPBA40 data
set \cite{shattuck2008}%
\footnote{
http://www.loni.ucla.edu/Atlases/LPBA40
}
publicly available from the Laboratory of Neuro Imaging at
the University of California, Los Angeles.  Preprocessing included
brain extraction using the provided masks and N4 bias correction
\cite{tustison2010}.

Batch registration was facilitated by the inclusion of the proposed
method in our ANTs interface to the new Insight Toolkit registration
framework called \verb#antsRegistration# which is detailed in 
\cite{tustison2012}.  An active development version is found in 
the ANTS Sourceforge repository.%
\footnote{
http://sourceforge.net/projects/advants/
}
A portion of the batch script used to produce the registrations is 
found in Listing \ref{listing:script}.  Each image was histogram-
matched prior to application of an initial affine transform using
the Demons metric.  The resulting transform was composed with the
proposed work using a neighborhood cross correlation similarity metric
\cite{avants2008} of radius = 4.  A multi-resolution strategy was 
employed where the initial B-spline velocity field mesh size was 
$24\times24\times12\times1$ which was doubled at each level.  
Other registration parameters can be gleaned from the Listing \ref{listing:script}.
Batch processing was performed on the linux cluster at the 
University of Virginia with each job allocated 4 nodes for
multi-threading purposes.  Each registration took approximately
3 hours.

\vspace{2mm}
\lstset{frame = htb,
        framerule = 0.25pt,
        float,
        fontadjust,
        backgroundcolor={\color{listlightgray}},
        basicstyle = {\ttfamily\scriptsize},
        keywordstyle = {\ttfamily\color{listkeyword}\textbf},
        identifierstyle = {\ttfamily},
        commentstyle = {\ttfamily\color{listcomment}\textit},
        stringstyle = {\ttfamily},
        showstringspaces = false,
        showtabs = false
        numbers = none,
        numbersep = 6pt,
        numberstyle={\ttfamily\color{listnumbers}},
        tabsize = 2,
        language=bash,
        floatplacement=!h,
        caption={Representative script used for the LPBA40 
        evaluation.},
        captionpos=b,
        label=listing:script
        }
\begin{lstlisting}
# Register the n4 corrected moving LPBA subject to the 
# n4 corrected fixed LPBA subject.

antsRegistration --dimensionality 3 \
                 --output ${prefix} \
                 --use-histogram-matching 1 \
                 --transform Affine[0.5] \                # affine stage
                 --metric Demons[${fixed_n4},${moving_n4},1,0,Regular,0.01] \ 
                 --iterations 100x100x100 \
                 --smoothing-sigmas 4.0x3.0x2.0 \
                 --shrink-factors 4x3x2 \
                 --transform tvdmffd[0.75,24x24x12x1,4] \ # tv dmffd stage
                 --metric CC[${fixed_n4},${moving_n4},1,4] \ 
                 --iterations 40x50x2 \
                 --smoothing-sigmas 1.0x0.5x0.0 \
                 --shrink-factors 3x2x1

# Apply the resulting transforms (affine + tvdmffd) to the 
# moving labels.

antsApplyTransforms --dimensionality 3 \
                    --input ${moving_labels} \
                    --reference-image ${fixed_n4} \
                    --output ${moving_warped_labels} \
                    --n NearestNeighbor \
                    --transform ${prefix}1Warp.nii.gz \
                    --transform ${prefix}0Affine.mat \
                    --default-value 0
\end{lstlisting}

Registration assessment used the average target overlap 
\cite{tustison2009a} for each of the 56 LPBA40 labeled regions.
The numerical values of this assessment are given in Table 1.
These values were projected onto subject 1 for display in Fig. 1
which can be compared to a similar display given in \cite{Klein2009}
with the caveat that the colormap is slightly different.

\begin{figure}
\centering
\begin{tabular}{c}
  \includegraphics[width=120mm]{Figures/leftHemisphere.png} \\
  \includegraphics[width=120mm]{Figures/rightHemisphere.png}
\end{tabular}
\caption{Average regional total overlap measures mapped onto Subject 1
of the LPBA40 data set.}
\label{fig:lpba40}
\end{figure}

\setlength{\tabcolsep}{4pt}

\begin{table}[ht]
\begin{minipage}[b]{0.5\linewidth}\centering
\begin{tabular}{lc}
{\bf Region} & {\bf Overlap} \\
\hline
L superior frontal gyrus & 0.802 \\
R superior frontal gyrus & 0.810 \\
L middle frontal gyrus & 0.790 \\
R middle frontal gyrus & 0.778 \\
L inferior frontal gyrus & 0.753 \\
R inferior frontal gyrus & 0.728 \\
L precentral gyrus & 0.687 \\
R precentral gyrus & 0.674 \\
L middle orbitofrontal gyrus & 0.687 \\
R middle orbitofrontal gyrus & 0.671 \\
L lateral orbitofrontal gyrus & 0.669 \\
R lateral orbitofrontal gyrus & 0.607 \\
L gyrus rectus & 0.654 \\
R gyrus rectus & 0.652 \\
L postcentral gyrus & 0.604 \\
R postcentral gyrus & 0.620 \\
L superior parietal gyrus & 0.709 \\
R superior parietal gyrus & 0.701 \\
L supramarginal gyrus & 0.619 \\
R supramarginal gyrus & 0.623 \\
L angular gyrus & 0.594 \\
R angular gyrus & 0.620 \\
L precuneus & 0.668 \\
R precuneus & 0.659 \\
L superior occipital gyrus & 0.548 \\
R superior occipital gyrus & 0.561 \\
L middle occipital gyrus & 0.671 \\
R middle occipital gyrus & 0.696 \\
\hline
\end{tabular}
\end{minipage}
\begin{minipage}[b]{0.5\linewidth}
\centering
\begin{tabular}{lc}
{\bf Region} & {\bf Overlap} \\
\hline
L inferior occipital gyrus & 0.621 \\
R inferior occipital gyrus & 0.637 \\
L cuneus & 0.622 \\
R cuneus & 0.624 \\
L superior temporal gyrus & 0.750 \\
R superior temporal gyrus & 0.737 \\
L middle temporal gyrus & 0.660 \\
R middle temporal gyrus & 0.644 \\
L inferior temporal gyrus & 0.662 \\
R inferior temporal gyrus & 0.674 \\
L parahippocampal gyrus & 0.711 \\
R parahippocampal gyrus & 0.707 \\
L lingual gyrus & 0.694 \\
R lingual gyrus & 0.681 \\
L fusiform gyrus & 0.703 \\
R fusiform gyrus & 0.719 \\
L insular cortex & 0.812 \\
R insular cortex & 0.768 \\
L cingulate gyrus & 0.691 \\
R cingulate gyrus & 0.676 \\
L caudate & 0.704 \\
R caudate & 0.717 \\
L putamen & 0.773 \\
R putamen & 0.761 \\
L hippocampus & 0.753 \\
R hippocampus & 0.742 \\
cerebellum & 0.935 \\
brainstem & 0.899 \\
\hline
\end{tabular}
\end{minipage}
\vspace{1.0mm}
\caption{Average target overlap values for cross-registration of the first 
20 subjects of the LPBA40 data set.  A visual depiction of these values is
given in Figure \ref{fig:lpba40}.
}
\end{table}
